Windowed least-squares model reduction for dynamical systems

### 2020-10-12 response_to_reviewers/response.tex

\address[a]{Sandia National Laboratories,  Livermore, CA}
\end{frontmatter}

%\maketitle
We would like to thank both of the reviewers for their helpful comments. We have done our best to address all concerns, and we feel that the comments have improved the quality of the manuscript. We have primarily improved our numerical results section, where we have (1) included studies of convergence with basis dimension, (2) included a parameterized example, and (3) increased the difficulty of our cavity flow experiment. In addition, we have worked on our analyses to try and address several of the concerns raised by the reviewers. Please see our responses below, in red.
\section{Reviewer 2}
The article proposes a windowed least-squares approach for model reduction of dynamical systems. The methodology is proposed as a generalization of existing approaches such as the least square Petrov-Galerkin (LSPG) method and the space-time LSPG method. The proposed methodology is clearly presented and related with existing and previously proposed approaches. The WLS methods is numerically implemented using two different approaches a direct one, which first discretizes then optimizes and an indirect one which first optimizes then discretizes. The article proposes also a theoretical analysis of the proposed methods. The paper is well written and easy to follows. All the methods are presented with a great level of details and results are convincing therefore I recommend publication after the raised issues are addressed.

\subsection{Major comments}
My biggest concern is about the numerical results section that in my opinion should be improved
\begin{enumerate}
\item Both numerical results should also show a convergence analysis of the methodology with respect to the dimension of the trial subspace used for the projection. Is the methodology sensitive with respect to the dimension of the trial subspace? Does the error decreases monotonically as I increase the number of reduced basis functions?

{\color{red} We have added a new sections to both of our numerical experiments to address this question. We additionally examine convergence in the new example we added. We show that the behavior of the method is as one would expect: increasing the basis dimension yields both lower errors and lower residuals. We observe that the error monotonically decreases for larger window sizes. For smaller window sizes we do not see monotonic convergence of the error; this result has been seen in the community for, e.g., the Galerkin method and LSPG. We observe similar behavior with the residuals.}

\item In order to have a better understanding of the time saving of the implemented ROM, in figure 8 the authors should report also the wall-clock times of the FOM. Also figure 10 should have a comparison also with the FOM wall-clock times.
{\color{red} We have added CPU timings to our second and third numerical experiments to address this concern. In our first numerical experiment, which serves as a reproducible benchmark, we do not employ hyper-reduction, and therefore do not obtain computational savings. Our reasoning for not including CPU timings in the first draft of the manuscript was that we felt that the timing with respect to the least-squares Petrov--Galerkin method was the relevant metric to report. The reason for this is that the speedup obtained by the ROM for nonlinear problems is highly dependent on the sample mesh, the associated stencil, and the computational kernels involved in evaluating residuals on the sample mesh. As LSPG has been shown to yield significant computational speedups in previous work (e.g.,~\cite{carlberg_lspg_v_galerkin}), and WLS with S-reduction subspaces is a windowed extension of the LSPG approach, we felt that comparing the speed of the proposed method to LSPG is the relevant timing. However, we do agree that comparison to the FOM walltime is a metric of interest, and therefore we have worked hard to integrate it into the second and third examples. For the problems considered, WLS gives between a 2x and a 20x speedup over the FOM, depending on the basis dimension and window size. We also investigate the performance of freezing the Jacobian in the Gauss--Newton algorithm on the solution speed.}

@@

\item I think that both numerical examples are rather simple to test the effectiveness of the method in real world situations. Moreover the numerical examples are used only to the test the reproductive ability of the ROM. I think that the authors should show the effectiveness of the methods on a more complex example which involves parametric dependency or time extrapolation.

{\color{red} To address this concern, we have (1) increased the difficulty of the second numerical experiment by raising the Mach number such that is supersonic, thus adding a complex shock boundary layer interaction to the problem, and (2) added a new numerical experiment, which starts at 6.3, with varying parameters. With respect to the new parameterized experiment, in addition to presenting various ROMs, we demonstrate that a linear surrogate model, in where we build a multivariate linear model for each degree of freedom of the dynamical system, fails to yield an adequate response. We appreciate the reviewers remarks, as we feel that the new numerical results provide stronger evidence of the utility of the approach.}

@@

\end{enumerate}

\subsection{Minor comments}

\begin{itemize}
\item page 25 second line, $\Delta T = .002 \rightarrow  \Delta T = 0.002$

{\color{red} Thank you, good catch!}

@@

\item Could the authors explain what is the meaning of the integrated residual norm for the l2
projection? This is not clear to me.

{\color{red} Thank you for drawing this to our attention. To clarify, we have (1) rephrased this as the ``time-integrated residual norm" and (2) added the clarifying statement" ``(i.e., the time-integrated norm of the residual evaluated at the $\elltwo$-orthogonal projection of the FOM solution onto the trial subspace)". In the previous manuscript, we also had an inconsistency of using both the term ``space--time" error norm and "integrated" error norm to refer to the same thing; we have corrected this in the revision by replacing space--time error norm with time-integrated norm.}

@@

{\color{red} Thank you, we have updated figures with relative errors.}
\end{itemize}

\section{Reviewer 3}
The authors presented a space-time model reduction method that aims to generalize/unify a number of different model reduction techniques and improve some shortcomings (exponential error growth in time, strong dependence on time discretization, non-intuitive relationship between error and time step size, etc). The proposed approach minimizes the time-continuous residual (integrated) over some time window in a space-only or space-time trial space. Two different approaches are proposed for solving the resulting WLS optimization problem: direct and indirect; both approaches perform similarly in numerical experiments. Numerical experiments show improved performance of WLS over the Galerkin and minimum-residual methods (called LSPG in this paper) and extensive testing is done to determine the performance of the method with respect to time step and time window. The proposed method is interesting and novel; however, I am recommending major revisions as there are
important aspects of the analysis and numerical experiments that should be addressed.

\subsection{Minor comments}
\begin{enumerate}
\item The "windowing" approach proposed in this work is very similar or identical to the concept of a space-time slab. There is no need to change the name of the method, but the similarity should be mentioned at least in the introduction and relevant literature cited.

{\color{red} We agree with the reviewer in this similarity (in fact, we used the terminology time-slab in our early drafts of the manuscript) , and have added several relevant citations. We are glad to add additional ones if the author has any recommendations. As to the difference between our approach and a time-slab, our perspective is as follows: While the term time-slab can entail many things, from our perspective, typically a space--time slab is viewed as a discretization technique. For example, in a space--time DG method, we might break the space--time domain down into space--time slabs, then represent the spatial and temporal dependence of the solution with polynomials on each space--time slab. In our setup, we break the space--time domain down into windows, and then rather than discretizing in time, we reformulate the problem as a least-squares problem on each window. We are then flexible to use any time discretization technique on that window, so perhaps a window may comprise various space--time slabs in the traditional sense. However, the two concepts are very closely tied and its easy to make a counter argument to our perspective! We have added the following to the manuscript, at the end of page 2 in or related work section:

@@

\textit{In this work, we refer to the “state-space” dimension of an ordinary differential equation as the “spatial” dimension. We use this notation due to the common application of projection-based ROMs to semi-discretized partial differential equations. We emphasize, however, that the presented approach is applicable to ordinary differential equations that do not associate with some type of spatial discretization, i.e., as in molecular dynamics.}

We have not modified the terminology in the paper, however. We feel that it is best to not do this as 1) previous work on which this builds uses the notion of space in a similar fashion (i.e., the space--time LSPG approach), and 2) we feel that referring to the ``spatial" dimension as, i.e., the ``state-space" dimension becomes cumbersome and slightly confusing. We are glad to adapt a new notation at the editors request, however.}

 \item  Assumption A4 is also very strong. What would happen if a similar assumption was introduced into the analysis of the Galerkin or LSPG methods? To provide a "fair" comparison between the error bound for the WLS method and Galerkin/LSPG methods (to support the claim the WLS method overcomes the exponential error growth) the authors are attempting to improve upon, the same (or very similar) assumptions should be made in both cases.

{\color{red} We appreciate this criticism, and agree with it. In our revision, we have removed this assumption by slightly modifying our analysis. We now place a bound on the in-plane error, and the constant multiplying the residual term now has an additional factor of two. The main take away of  the analysis remains the same, in that the error grows recursively with respect to the number of windows, opposed to the number of time steps.}

@@

We are in complete agreement, however, with the fact that computing these Lipshitz constants is not practical. Doing so would be extremely difficult, if not impossible. We have added a remark to our analysis about this. 
}

\item I appreciate the authors extensive numerical experiments to study the performance/cost of the method, particularly with dt and the time window; however, there is some important information missing. For Section 6.1, the authors freeze the basis size and perform dt and dT sweeps, but provide no insight on how the various methods perform (accuracy and cost) as the basis size increases. Since the WLS approach is more expensive than Galerkin/LSPG for a given basis size, this basis size sweep will provide critical information regarding the relative performance (accuracy and cost) of the methods. To avoid running an exponential number of cases, the optimal dt and dT determined from the previous studies can be used. Similarly for Section 6.2: two basis sizes were considered but the energy only varies from 95\% to 97\%; many papers have shown higher energy (99\% or more) is required for complex problems like this, which should be done for a "fair" comparison to Galerkin/LSPG.

{\color{red} We appreciate these comments. In an effort to address them, we have added new sections to all of our numerical experiments where we assess the performance of the method as a function of basis dimension. For the first numerical experiment, we do this for numerous window sizes, while for the second experiment we do it for a single window size. We do it for numerous window sizes for our third experiment. We additionally report CPU timings for all of these different permutations. With respect to the final comment on basis dimension, we had a typo in reporting the energy content, thank you for pointing this out. This has been corrected in the revised manuscript, and all bases considered contain $99.99\%$ of the energy.}

\item Neither of the examples vary parameters (purely reproducing the solution). While this was explicitly stated in the paper as not a priority, it significantly lessens relevance of the paper. I strongly recommend including a parametric example.
{\color{red} To address this concern, we added a new numerical experiment with varying parameters.  In this new experiment, we explore the performance of LSPG and WLS for prediction at a parameter outside the training set. We study the impact of window size and basis dimension on the results. In addition to presenting various ROMs, we demonstrate that a linear surrogate model, in where we build a multivariate linear model for each degree of freedom of the dynamical system, fails to yield an adequate response. We observe WLS to yield improvements over the LSPG approach.}

### 2020-10-12 cavity.tex
latest

The reduced-order models leverage POD to construct the \spatialAcronym\ trial basis and use q-sampling~\cite{qdeim_drmac} based on snapshots of the FOM solution to select the sampling points (and as a result the weighting 
matrix $\stweightingMatArg{}$); we use a constant weighting matrix across all windows, e.g., $\stweightingMatArg{n} \equiv \stweightingMatArg{}$, $n=1,\ldots,\nslabs$. The process used to construct the initial conditions, trial subspaces, and weighting matrices for the ROMs is as follows:
\begin{enumerate}
\item Initialize the FOM with uniform free-stream conditions.
\item Evolve the FOM until a statistically steady state has been reached.
\item Reset the time coordinate, $0 \leftarrow t$, and execute Algorithm~\ref{alg:pod} with $\stateInterceptArg{} = \bz, N_{\text{skip}} = 8$, $K = \{50,100,156,200\}$, and $\podIP =  \mass_{\text{DG}}$ over $t \in [0,50]$ to construct the four trial subspaces tabulated in Table~\ref{tab:rom_basis_details}. 
\item Execute Algorithm~\ref{alg:qdeim} with $N_{\text{skip}} = 8$, $n_s = 250$ to obtain the sampling point matrix of dimension $\stweightingMatOneArg{} \in \{0,1\}^{10170 \times \fomdim }$. 
\end{enumerate}
Figure~\ref{fig:cav_sampmesh} shows the resulting sample mesh used in the ROM simulations. To depict the nature of the flow, Figure~\ref{fig:fom_sols_cav} shows snapshots of the vorticity and density gradient fields generated by the FOM for several time instances used in training.  
\begin{table}[]
\begin{centering}
\begin{tabular}{c c c c}
\hline
Basis \# & Trial Basis Dimension ($K$) &  Energy Criterion & Sample Points ($n_S$) \\
\hline
1    & $50$ &  $99.99738\%$ & $10170$ \\
2    & $100$ & $99.99956\%$ & $10170$ \\
3    & $156$ & $99.99990\%$ & $10170$ \\
4    & $200$ & $99.99999\%$ & $10170$ \\
\hline
\end{tabular}
\caption{Summary of the various basis sizes employed in the cavity flow example.}
\label{tab:rom_basis_details}
\end{centering}
\end{table}

@@

\item \textit{LSPG ROM with collocation}: We consider a collocated LSPG ROM, which is built on top of the FOM discretization using the CN scheme for temporal 
discretization. We employ a time step of $\Delta t  = 0.05$. The implementation is the same as previously described.

\item \textit{\methodAcronymROMs\ with collocation}: We consider \methodAcronymROMs\ solved via the direct method. The ROMs use the CN time discretization with a time step of 
$\Delta t = 0.05$. The implementation is the same as previously described. 
%\begin{itemize}
%\item Direct Method: \methodAcronym ROMs solved via the direct method are again implemented with the same Crank--Nicolson time discretization with a nominal 
%time-step size of $\Delta t = 0.1$. 
%\end{itemize}
\end{itemize}

@@

Next, Figure~\ref{fig:cav_wallclock}
shows the wall-clock times\footnote{We computed wall-clock times over the time inteval $t \in [0,1]$} of the \methodAcronymROMs\ as compared to (1) the LSPG ROM (Figure~\ref{fig:cav_wallclock_lspg}) and (2) the FOM (Figure~\ref{fig:cav_wallclock_fom}).  As expected, increasing $\Delta T$ again leads to an increase in computational cost; minimizing the residual over a window comprising 20 time instances yields a 4x increase in cost over LSPG. With respect to the FOM, we find that WLS ROMs with window sizes of $\Delta T = 0.1, 0.25, 0.5,$  and $1.0$ are approximately $25$x, $16$x, $12$x, and $8$x faster, repectively. The LSPG ROM is approximately 30x faster than the FOM.  We make several remarks about these CPU times. First, it must be noted that, due to a decreased stiffness in the system, we are able to run the ROMs at a time step that is 10x greater than the FOM. Second, we note that in our DG code, the vast majority of CPU time in the WLS ROMs is spent computing the Jacobians. For WLS ROMs emplying a sample mesh of size $n_s=10170$ with $K=156$ basis vectors and a window size of $\Delta T = 0.025$, it takes approximately $2$(s) to compute the residual over one time window; it takes $500$(s) to compute the Jacobian via automatic differentiation. Our full-order model solver uses a Jacobian-free GMRES solver, and thus is not subject to this bottleneck. While this is in part due to the fact that high-order DG methods have dense block Jacobians (each element has 45 DOFs associated with it for a $p=2$ discretization of the Navier--Stokes equations), we expect that significant computational speedups can be obtained by employing, e.g., analytic Jacobians, matrix-free Jacobian transpose methods~\cite{doi:10.2514/6.2016-0833}. This will be a topic of future work. Nonetheless, the present example demonstrates some of the practical challenges that exist for least-squares ROMs such as WLS and LSPG.

>>>>

The reduced-order models leverage POD to construct the \spatialAcronym\ trial basis and use q-sampling~\cite{qdeim_drmac} based on snapshots of the FOM solution to select the sampling points (and as a result the weighting 
matrix $\stweightingMatArg{}$); we use a constant weighting matrix across all windows, e.g., $\stweightingMatArg{n} \equiv \stweightingMatArg{}$, $n=1,\ldots,\nslabs$. The process used to construct the initial conditions, trial subspaces, and weighting matrices for the ROMs is as follows:
\begin{enumerate}
\item Initialize the FOM with uniform free-stream conditions.
\item Evolve the FOM until a statistically steady state has been reached.
\item Reset the time coordinate, $0 \leftarrow t$, and execute Algorithm~\ref{alg:pod} with $\stateInterceptArg{} = \bz, N_{\text{skip}} = 8$, $K = \{50,100,156,200\}$, and $\podIP =  \mass_{\text{DG}}$ over $t \in [0,50]$ to construct the four trial subspaces tabulated in Table~\ref{tab:rom_basis_details}. 
\item Execute Algorithm~\ref{alg:qdeim} with $N_{\text{skip}} = 8$, $n_s = 250$ to obtain the sampling point matrix of dimension $\mathbf{S} \in \{0,1\}^{10170 \times \fomdim }$.
\item Set $\stweightingMatOneArg{} = \mathbf{S} \mass_{\text{DG}}$. 
\end{enumerate}
Figure~\ref{fig:cav_sampmesh} shows the resulting sample mesh used in the ROM simulations. To depict the nature of the flow, Figure~\ref{fig:fom_sols_cav} shows snapshots of the vorticity and density gradient fields generated by the FOM for several time instances used in training.  
\begin{table}[]
\begin{centering}
\begin{tabular}{c c c c}
\hline
Basis \# & Trial Basis Dimension ($K$) &  Energy Criterion & Sample Points ($n_S$) \\
\hline
1    & $50$ &  $99.99738\%$ & $10170$ \\
2    & $100$ & $99.99956\%$ & $10170$ \\
3    & $156$ & $99.99990\%$ & $10170$ \\
4    & $200$ & $99.99999\%$ & $10170$ \\
\hline
\end{tabular}
\caption{Summary of the various basis sizes employed in the cavity flow example.}
\label{tab:rom_basis_details}
\end{centering}
\end{table}

@@

\item \textit{LSPG ROM with collocation}: We consider a collocated LSPG ROM, which is built on top of the FOM discretization using the CN scheme for temporal 
discretization. We employ a time step of $\Delta t  = 0.05$. The implementation is the same as previously described in Section~\ref{sec:sod_rom_description}.

\item \textit{\methodAcronymROMs\ with collocation}: We consider \methodAcronymROMs\ solved via the direct method. The ROMs use the CN time discretization with a time step of 
$\Delta t = 0.05$. The implementation is the same as previously described in Section~\ref{sec:sod_rom_description}. 
\end{itemize}

@@

Next, Figure~\ref{fig:cav_wallclock}
shows the wall-clock times\footnote{We computed wall-clock times over the time interval $t \in [0,1]$} of the \methodAcronymROMs\ as compared to (1) the LSPG ROM (Figure~\ref{fig:cav_wallclock_lspg}) and (2) the FOM (Figure~\ref{fig:cav_wallclock_fom}).  As expected, increasing $\Delta T$ again leads to an increase in computational cost; minimizing the residual over a window comprising 20 time instances yields a 4x increase in cost over LSPG. With respect to the FOM, we find that WLS ROMs with window sizes of $\Delta T = 0.1, 0.25, 0.5,$  and $1.0$ are approximately $25$x, $16$x, $12$x, and $8$x faster, respectively. The LSPG ROM is approximately 30x faster than the FOM.  We make several remarks about these CPU times. First, it must be noted that, due to a decreased stiffness in the system, we are able to run the ROMs at a time step that is 10x greater than the FOM. Second, we note that in our DG code, the vast majority of CPU time in the WLS ROMs is spent computing the Jacobians. For WLS ROMs employing a sample mesh of size $n_s=10170$ with $K=156$ basis vectors and a window size of $\Delta T = 0.025$, it takes approximately $2$(s) to compute the residual over one time window; it takes $500$(s) to compute the Jacobian via automatic differentiation. Our full-order model solver uses a Jacobian-free GMRES solver, and thus is not subject to this bottleneck. While this is in part due to the fact that high-order DG methods have dense block Jacobians (each element has 45 DOFs associated with it for a $p=2$ discretization of the Navier--Stokes equations), we expect that significant computational speedups can be obtained by employing, e.g., frozen Jacobians, analytic Jacobians, matrix-free Jacobian transpose methods~\cite{doi:10.2514/6.2016-0833}. We briefly investigate the use of frozen Jacobians in the following numerical experiment and Appendix~\ref{appendix:gnupdate}, where we show that employing a Gauss--Newton algorithm with frozen Jacobians can yield substantial speedups. Nonetheless, the present example demonstrates some of the practical challenges that exist for least-squares ROMs such as WLS and LSPG.
